---
title: "chapter4.Rmd"
author: "Xindi Huang"
date: "2023-11-27"
output: html_document
---


# Assignment 4: Clustering and classification

```{r}
date()

# laod packages
library(tidyverse)
library(dplyr)
library(MASS)
if (!require(caret)) {
  install.packages("caret")
  library(caret)
}
```

## Boston dataset
```{r}
# Load the Boston data from the MASS package
data(Boston)

# Explore the structure and dimensions of the data
str(Boston)
dim(Boston)

```

The `Boston` dataset contains information on various features related to housing in different suburbs of Boston. The dataset has 506 observations and 14 variables.

Variables info :
- crim: Per capita crime rate by town.
- zn: Proportion of residential land zoned for lots over 25,000 sq.ft.
- indus: Proportion of non-retail business acres per town.
- chas: Charles River dummy variable (1 if tract bounds river, 0 otherwise).
- nox: Nitrogen oxides concentration (parts per 10 million).
- rm: Average number of rooms per dwelling.
- age: Proportion of owner-occupied units built prior to 1940.
- dis: Weighted mean of distances to five Boston employment centers.
- rad: Index of accessibility to radial highways.
- tax: Full-value property-tax rate per $10,000.
- ptratio: Pupil-teacher ratio by town.
- black: Proportion of blacks by town.
- lstat: Lower status of the population (percent).
- medv: Median value of owner-occupied homes in $1000s.

## Graphical overview and summaries of the variables

```{r}
# Graphical overview of the data
pairs(Boston)
# Summaries of the variables
summary(Boston)
```

### 1. Statistical summaries of the variables in the `Boston` dataset.

- **`crim` (Crime Rate):**
Indicats a right-skewed distribution.The majority of towns have lower crime rates, with 25% having a rate below 0.08205, but a few towns have high crime rate.

- **`zn` (Proportion of Residential Land):**
Shows a right-skewed distribution with many towns having little residential land.

- **`indus` (Proportion of Non-Retail Business Acres):**
Display a mix of industrial and residential areas.

- **`chas` (Charles River Dummy Variable):**
Most suburbs don't border the Charles river, only 6.92% of towns border the river.

- **`nox` (Nitrogen Oxides Concentration):**
Concentration ranges from 0.385 to 0.871. Mean of 0.5547 suggests a moderate level of nitrogen oxides in most zones, but some zones hace high level of nitrogen oxides.

- **`rm` (Average Number of Rooms):**
Display a normal distribution, shoing the most have median number of 6.208 rooms.

- **`age` (Proportion of Older Units):**
Mean of 68.57 suggests a mix of older and newer units. The distribution is left-skewed indicating many houses are older.

- **`dis` (Weighted Mean of Distances to Employment Centers):**
The right-skewed distribution of the `dis` variable suggests that most suburbs have a relatively short distance to employment centers, as indicated by lower values.

- **`rad` (Index of Accessibility to Radial Highways):**
`rad` variable appears bimodal, suggesting that suburbs tend to fall into two categories: those that are very close to highways and those that are very far

- **`tax` (Property Tax Rate):**
Shows bimodal, ranges from 187 to 711.

- **`ptratio` (Pupil-Teacher Ratio):**
Mean of 18.46 indicates a moderate pupil-teacher ratio.

- **`black` (Proportion of Blacks):**
Mean of 356.67 suggests a mix of racial proportions.

- **`lstat` (Lower Status of the Population):**
Mean of 12.65 indicates a mix of lower and higher-status populations.

- **`medv` (Median Value of Homes):**
Mean of 22.53 indicates a moderate median home value.

### 2. Relationships between the variables 
**Positive correlation:**
- `rm` and `medv`: as more rooms may be associated with higher home values.
- `idus` and `nox`: areas with more industrial land use might have higher concentrations of nitrogen oxides.
- `age` and `nox`: Older buildings might be associated with higher concentrations of nitrogen oxides.
- `rad` and `tax`: There could be a positive correlation between the accessibility to highways and the property tax rate.

**Negtive correlation**
- `dis` and `nox`: as areas closer to employment centers might have lower concentrations of nitric oxides.
- `rm` and  `lstat`: higher-status neighborhoods might have more rooms on average.
- `lstat` and `medv`: areas with a lower socioeconomic status might have lower home values.

### 3. Overall Observations:

- The dataset captures a diverse range of features related to housing in Boston suburbs.
- Many variables exhibit right-skewed distributions, suggesting that the majority of towns have lower values.

## Standardize the dataset and print out summaries of the scaled data

```{r}
# center and standardize variables
boston_scaled <- as.data.frame(scale(Boston))

# summaries of the scaled variables
summary(boston_scaled)

```

```{r}

# create a quantile vector of crim
boston_scaled$crim <- as.numeric(boston_scaled$crim)
bins <- quantile(boston_scaled$crim)
print(bins)

# creat a categorical variable 'crime' based on quantiles
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE)
summary(crime)

# create a string vector for labels
labels <- c("low", "med_low", "med_high", "high")

# adjust the cut function by adding the label argument
crime <- cut(boston_scaled$crim, breaks = bins, labels = labels, include.lowest = TRUE)

# look at the table of the new factor crime
table(crime)

# add the new categorical value in the original Boston dataset
Boston$crime <- crime
Boston$crime <- as.factor(Boston$crime)

# remove original crim from the dataset
Boston <- dplyr::select(Boston, -crim)

# divide the dataset into train and test sets (80% train, 20% test)
ind <- sample(nrow(Boston),  size = nrow(Boston) * 0.8)
train <- Boston[ind,]
test <- Boston[-ind,]

```


## Linear Discriminant Analysis 
```{r}
# Fit the linear discriminant analysis on the train set
lda_fit <- lda(crime ~ ., data=train)

# Summary the lda fitting
lda_fit

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda_fit, dimen = 2, col = classes, pch = classes) 

```

## Predictions and Cross-tabulation
```{r}
# Save the crime categories from the test set
correct_classes <- test$crime

# Remove the categorical crime variable from the test dataset
test <- dplyr::select(test, -crime)

# Predict the classes with the LDA model on the test data
lda_pred <- predict(lda_fit, newdata = test)$class

# Cross-tabulate the results with the crime categories from the test set
table(correct_classes, lda_pred)

```

The results of the Linear Discriminant Analysis (LDA) in predicting crime categories on the test set:

- **Accurate Predictions:** The diagonal entries showcase instances where the model correctly identified the crime category. For instance, it accurately predicted 17 cases in the "low" category, 13 cases in "med_low," 9 cases in "med_high," and 24 cases in the "high" category.

- **Misclassifications:** Off-diagonal entries reveal instances where the model made incorrect predictions. For example, in the "low" category, there were 12 instances misclassified as "med_low," and 1 instance as "med_high." Notably, there were no instances misclassified as "high" in this category.

- **Category-specific Performance:** Each row corresponds to the actual crime category, allowing us to delve into how well the model performed within each specific category.

In essence, this table offers a granular view of the model's strengths and areas for improvement across different crime categories. It provides valuable insights for refining the model and enhancing its accuracy in predicting crime rates.

## Clustering

```{r}
# Reload the Boston dataset
data(Boston)

# Standardize the dataset
boston_scaled <- scale(Boston)

# Calculate distances between observations
dist_eu <- dist(boston_scaled)
summary(dist_eu)
```

## k-means algorithm with varying numbers of clusters (k)

```{r}
# Determine the number of clusters
k_max <- 10

# Calculate total within-cluster sum of squares (TWSS) for different cluster numbers
twss <- sapply(1:k_max, function(k) kmeans(boston_scaled, centers = k)$tot.withinss)

# Visualize the TWSS
plot(1:k_max, twss, type = "b", pch = 19, frame = FALSE, col = "blue", xlab = "Number of clusters", ylab = "Total Within Sum of Squares")
```

```{r}
# Choose the optimal number of clusters
optimal_clusters <- 4

# Run k-means with the optimal number of clusters
kmeans_result <- kmeans(boston_scaled, centers = optimal_clusters)

# Creating a data frame for plotting
plot_data <- as.data.frame(boston_scaled)
plot_data$cluster <- factor(kmeans_result$cluster)

# Visualize clusters using pairs() with colors
ggplot(plot_data, aes(x = rm, y = lstat)) +
  geom_point(aes(color = cluster)) +
  labs(color = 'Cluster') 
```

**Cluster Distribution:** The plot shows how the observations are grouped into four different clusters. Each cluster is represented by a different color.

**Cluster 1:** Higher `rm` and moderate to low `lstat` values, indicating larger houses with a lower proportion of lower-status population.

**Cluster 2:** Moderate to high rm values but very low lstat values, suggesting these areas have larger houses and very low lower-status population proportions.

**Cluster 3:** Reflects neighborhoods with smaller homes (low `rm``) where the community leans towards a higher percentage of lower-status residents.

**Cluster 4:** Moderate `rm` values and `lstat` values, indicating that the average-sized rooms with a moderate lower-status population.


**Correlation Inference:** The negative correlation between the number of rooms (`rm`) and the percentage of lower-status population (`lstat`) aligns with our expectations. In areas with more rooms, there tends to be a lower percentage of lower-status residents.

**Cluster Separation:** The separation between clusters indicates how distinct the groups are based on the two variables used. Cluster 1 stands out from the rest, indicating that areas with small houses and very low lower-status proportions are distinct from the others.


```{r}
# Standardize the original Boston dataset
original_boston_scaled <- as.data.frame(scale(Boston))

# Perform k-means with some reasonable number of clusters
reasonable_clusters <- 4
kmeans_original <- kmeans(original_boston_scaled, centers = reasonable_clusters)

# Perform LDA using the clusters as target classes
lda_fit <- lda(kmeans_original$cluster ~ ., data = original_boston_scaled)

# Visualize LDA results with a biplot
plot(lda_fit)
```
```{r}
lda_fit$scaling
```

The values in the output represent the coefficients or loadings of each original variable on the linear discriminants (LD1, LD2, and LD3) in the Linear Discriminant Analysis (LDA) model:

- **LD1, LD2, LD3:** These are the linear discriminants generated by the LDA. Each row corresponds to a variable in the original dataset.

- **Coefficients (Loadings):** These values indicate the contribution of each variable to the respective discriminant. A positive coefficient means that an increase in the variable is associated with a higher value of the discriminant, while a negative coefficient suggests the opposite.

- **For LD1:**
  - `crim`: 0.01913416 (positive) - A higher crime rate is associated with a higher value of LD1.
  - `zn`: 0.15673395 (positive) - Higher residential land zoned for large lots is associated with a higher value of LD1.
  - `indus`: -0.52081410 (negative) - Higher industrial acres per town is associated with a lower value of LD1.
  
These coefficients indicate the direction and strength of the relationship between the original variables and the discriminants.
